---
title: "Compound half-life analysis"
output: html_document
---

### Goal and Approach

We want to determine the elimination half-lives of the molecules in `pk_data.csv`.
First, I will get a sense of what the data looks like and produce a series of exploratory plots.
Then, once I know what kind of problems are in the data, I will exclude outliers.
Finally, I will do some non-linear modeling to determine the elimination half-lives of the compounds.

### Setup

```{r}
library(ggplot2)
library(tidyverse)

data <- read.csv("pk_data.csv") |>
  mutate(
    is_control = is_control == "True"
  )
```

### Exploratory Analysis

```{r}
data |>
  mutate(
    hours_post_dose = if_else(
      is.na(hours_post_dose),
      0, hours_post_dose
    )
  ) |>
  filter(
    is_control
  ) |>
  ggplot() +
    geom_point(
      aes(
        x = hours_post_dose,
        y = concentration_ng_ml,
        colour = molecule_id
      ),
      alpha = .75
    ) +
    scale_y_continuous(transform = "log10") +
    theme_bw()
```

The above plot lets me know that long lasting compounds will have a 2 phase decay curve.
The first phase is distribution with a small amount of elimination before a major elimination phase comes.
It is likely that distribution rate is very similar between all of the compounds and is over by hour 20 or so as evidence in CTRL_long.
The difference in shape for the distribution phase of CTRL_medium and CTRL_short is likely coming from elimination that is a larger component of the effect.
Next, I will get a sense of the general distribution of curves.

```{r}
data |>
  mutate(
    hours_post_dose = if_else(
      is.na(hours_post_dose),
      0, hours_post_dose
    )
  ) |>
  group_by(molecule_id, hours_post_dose, is_control) |>
  summarise(
    concentration_ng_ml = median(concentration_ng_ml),
    .groups = "drop"
  ) |>
  ggplot() +
    geom_line(
      aes(
        x = hours_post_dose,
        y = concentration_ng_ml,
        group = molecule_id,
        color = is_control
      ),
      alpha = .5
    ) +
    scale_y_continuous(transform = "log10") +
    theme_bw()
```

Some things I notice here. One of which is that, after their initial spike, some compounds have a second small spike which may mess up non-linear modeling.
Furthermore, most compounds do not fully eliminate within the experiment, including the fast eliminating compounds.
For this latter group, the effect is likely technical, but it is still good to note since it will need to be taken into account when we do calculations.

Lets look at a distribution of when the median of each time point falls bellow 50% of the 1hr concentration, which is a crude order of half-life
```{r}
threshold_point <- data |>
  filter(
    !is.na(hours_post_dose) & hours_post_dose >= 1
  ) |>
  group_by(
    molecule_id, hours_post_dose, is_control
  ) |>
  summarise(
    concentration_ng_ml = median(concentration_ng_ml),
    .groups = "drop"
  ) |>
  group_by(molecule_id) |>
  mutate(
    passed_threshold = concentration_ng_ml < .5 * first(concentration_ng_ml)
  ) |>
  ungroup() |>
  filter(passed_threshold) |>
  distinct(molecule_id, .keep_all = T)

threshold_point|>
  ggplot() +
    geom_bar(
      aes(
        x = hours_post_dose
      )
    ) +
    geom_text(
      aes(
        x = 50,
        y = 45,
        label = paste("Total Peptides:", dim(threshold_point)[1])
      ),
      hjust = 1,
      data = data.frame()
    ) +
    theme_bw()
```